---
title: "Getting Started with the ABM Package in R"
author: "Sasha D. Hafner and Frederik R. Dalby"
date: "`r Sys.Date()`"
fig_width: 6 
fig_height: 4 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ABM_start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=4)
options(width = 85)
```

# 1. Overview
The ABM model predicts conversion of organic material in animal manure or other (high-moisture) organic wastes to methane (CH~4~), carbon dioxide (CO~2~) and ammonia (NH~3~) under anaerobic conditions.
The name comes from **a**naerobic **b**iodegradation **m**odel.
With multiple methanogen groups and group-specific parameters describing kinetics and yield, the model can predict realistic short- and long-term responses to temperature change and other perturbations.
Although it was prediction of CH~4~ emission from stored animal slurry (liquid manure) in unheated channels or tanks that prompted development of the model, the model can be used to simulate CH~4~ emission or biogas production from other organic waste under a range of conditions, including in anaerobic digesters.
The purpose of this document is to demonstrate the use of the ABM R package, which is a flexible implementation of the model.
Details on the model will be available in publications in the near future.

# 2. Installation
The ABM package is available on GitHub and so can be installed with the `install_github()` function from the devtools package, which must be installed first.
These steps should be carried out once to install both packages:

```{r, eval=FALSE}
install.packages('devtools')
devtools::install_github('AU-BCE-EE/ABM', build_vignettes = FALSE)
```

And to use the ABM model, the package must be loaded.

```{r,eval= TRUE}
library(dplyr)
library(ABM)
```

And to view this vignette, use:

```{r}
vignette('ABM_start')
```

```{r,eval = FALSE, echo=FALSE}
ff <- list.files('../R', full.names = TRUE)
for (i in ff) source(i)
```

# 3. A demonstration
A demonstration is presented in this section to show what the abm() function can do.
For a more incremental introduction to the function, see the following sections.

Let's assume a cattle farm produces manure at a rate of 10000 kg (fresh mass) per day.
Manure composition is typical, with volatile solids (VS) ca. of 11% of total mass, and degradability matching default values.
In our example, manure is transferred to an outdoor storage tank every day, and we can ignore any minor emission that occurs from the small quantity of fresh manure in the barn.
Further, let's assume we have measurements that show the temperature of stored manure ranges from a low of 3&degC in January to 22&degC in August, following the values shown below. 

```{r}
temp_dat <- data.frame(time = c(1, 50, 190, 240, 290, 315, 365), temp_C = c(8, 9, 26, 28, 26, 9, 9)) 
plot(temp_C ~ time, data = temp_dat, type = 'l')
```

The slurry tank is emptied twice per year, in March and September.
Considering that upon "emptying", 0.2 m of slurry is left in the tank, the storage height is 4 m and the surface area of the storage tank is 1000 m2. We assume that we start with the tank being 30% full. 

We'll duplicate the temperature data so we can simulate multiple years.

```{r}
t3 <- t2 <- temp_dat
t2$time <- t2$time  + 365
t3$time <- t3$time + 730
temp_dat <- rbind(temp_dat, t2, t3)
```



```{r}
out0 <- abm(3*365, add_pars = list(temp_C = temp_dat, slurry_prod_rate = 10000, storage_depth = 4, 
                                   area = 1000, empty_int = 182.5, 
                                   slurry_mass = 0.3 * 1000 * 4))
```
One could also have speficied a constant temperature by replacing "temp_C = temp_dat" with "temp_C = 20" in the "add_pars" argument. 

Slurry mass increases at a constant rate until it is emptied after 182.5 days.
In this simulation, 0.2 m is assumed to be left behind. A warning will be given manure production and empty intervals result in overfilling of the storage tank. 
Temperature exactly follows the curve shown above (with interpolation as needed).

```{r}
plot(slurry_mass/1000 ~ time, data = out0, type = 'l', xlab = 'Time (d)', ylab = 'Slurry mass (t)')
plot(temp_C ~ time, data = out0, type = 'l', xlab = 'Time (d)', ylab = 'Slurry mass (t)')
```

What is the predicted response of methanogen populations?

```{r}
matplot(out0$time, out0[, nn <- c('m0','m1','m2')]/1000, 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Microbial biomass (kg)')
legend('topleft', nn, col = 1:6, lty = 1)
```

We see a shift away from the most cold-tolerant (`m0` and `m1`, 18&degC optimum) to groups with higher optima (especially `m2`, 28&degC optimum) as the slurry warms, and the reverse during cooling.

Methane production depends on these populations, but also available substrate and hydrolysis rate (itself temperature-dependent).
Particulate substrate, in turn, depends on the quanity of slurry in the storage, but also prior consumption.
Together these effects give the complex pattern in CH~4~ emission (rate and then cumulative) shown below.

```{r}
plot(CH4_emis_rate ~ time, data = out0, type = 'l', xlab = 'Time (d)', ylab = 'Emission rate (g/d)')
plot(CH4_emis_cum ~ time, data = out0, type = 'l', xlab = 'Time (d)', ylab = 'Cumulative emission (g)')
```
Some of the spikes in CH~4~ production are related to low methanogen activity, which is reflected in VFA accumulation.
Temperature plays a major role in determining this pattern.

```{r}
plot(VFA_conc ~ time, data = out0, type = 'l', xlab = 'Time (d)', ylab = 'VFA conc. (g/kg)')
```

Because all output variables seem stable after two years, we might exclude the first 730 days when, for example, evaluating mitigation options.
Furthermore, we can also focus on total emission in the last year.

```{r}
out0a <- abm(3*365, add_pars = list(temp_C = temp_dat, storage_depth = 4, area = 1000, slurry_mass = 0.3 * 4 * 1000, empty_int = 182.5), startup = 730, value = 'summ')
out0a
```

Results show total annual emission of about 10 t CH~4~, representing 31% of available COD.

```{r}
out0a['CH4_emis_cum']
out0a['f_COD_CH4']
```

How much might acidification reduce emission?
We can easily make a comparison to a scenario with pH reduced to 5.5, with everything else the same.pH is indirectly affecting methane because reduced pH increase concentration of toxic H2S, and protonated VFAs.


```{r}
out0b <- abm(3*365, add_pars = list(temp_C = temp_dat, storage_depth = 4, area = 1000, slurry_mass = 0.3 * 4 * 1000, empty_int = 182.5, pH = 5.5), startup = 730, value = 'summ')
out0b
```

The predicted reduction is 98.6%.

```{r}
1 - out0b['CH4_emis_cum'] / out0a['CH4_emis_cum']
```

# 4. A simple example: methane emission from stored slurry
By default, the `abm()` function simulates degradation of animal manure accumulating in a 33 m^3^ storage tank or channel with a 30 day emptying interval.
Fresh slurry is added continuously at a rate of 1000 kg d^-1^, and when emptied a residual of 10% of the total manure mass is left in the storage.
Default values are included for all arguments, including the first two, which set the length of the simulation (365 d) and the time interval in the output (1 d).

In this example, the model is used to predict dynamics of CH~4~ emission, microbial biomass, and VFA accumulation.
The following call runs the ABM model with default argument values.

```{r}
out1 <- abm()
```

Output is, by default, a data frame with predicted variables over time (see Section 5 for alternatives).
Typically the primary variable of interest is CH~4~ emission, which is returned as a total (g) and rate (g/d), overall or normalized to COD or VS loading.
The complete list of these output variables are:

```{r}
names(out1[grepl('^CH4', names(out1))])
```

Total cumulative emission (g) and emission rate (g/d) are plotted below.

```{r}
plot(CH4_emis_cum ~ time, data = out1, type = 'l', ylab = 'Cumulative emission (g)')
plot(CH4_emis_rate ~ time, data = out1, type = 'l', ylab = 'Emission rate (g/d)')
```

Microbial biomass (g) is given in columns with names that match those used for the names of the groups (defaults shown below, set within the `grp_pars` argument--see Section 6).

```{r}
matplot(out1$time, out1[, nn <- c('m1', 'm2', 'm3', 'm4', 'm5')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Microbial biomass (g)')
legend('topleft', nn, col = 1:6, lty = 1)
```

Because of a default temperature of 23&deg (NTS: why so high?) methanogen `m2` dominates under default conditions.
Biomass concentrations (g per kg of slurry) may be more informative.

```{r}
plot(m3_conc ~ time, data = out1, type = 'l')
```

Dynamics in production of CH~4~ are often related to VFA accumulation, and VFA mass (g) and concentration (g/kg) can be extracted.

```{r}
plot(VFA ~ time, data = out1, type = 'l')
plot(VFA_conc ~ time, data = out1, type = 'l')
```
For more information on the many output variables returned by `abm()`, see the section on that topic below.

# 5. Setting parameter values
Although the ABM model is relatively simple, explicitly simulating the activity of multiple microbial groups means there are many parameters. 
The complete list and definitions can be seen in the help file, accessible with the following command.

```{r,eval=FALSE}
?abm
```

Alternatively, use `args()` just to see the arguments and default values.

```{r}
args(abm)
```

Parameters are grouped to make changes easier (and to prevent input mistakes) and to limit the number of parameter names that are needed.
The `mng_pars` argument contains parameters related to management; `man_pars` describes the incoming manure or feed; `grp_pars`, the most extensive argument, is used to define the microbial groups; `mic_pars` contains other microbial parameters that do not vary among groups; and `chem_pars` sets some chemical/physical parameters.
But there are also some built-in shortcuts to make small tweaks simple.
In particular, the `add_pars` argument makes life easy.

As an example, the composition of the fresh slurry (influent, or feed) is set with the `man_pars` argument, which is a list of solute concentrations and pH.
By default:
```
man_pars = list(conc_fresh = list(S2 = 0.0, SO4 = 0.2, TAN = 1.0, 
                                  VFA = 4.0, Sp = 65, COD = 170), 
                pH = 7), ...
```
To simulate a lower pH then, the following call could be used:

```{r}
out2 <- abm(365, 1, man_pars = list(conc_fresh = list(S2 = 0.0, SO4 = 0.2, TAN = 1.0, 
                                                       VFA = 4.2, Sp = 65, COD = 160), 
                                     pH = 6))
```

Below CH~4~ emission rate is compared to the default predictions.

```{r}
plot(CH4_emis_rate ~ time, data = out1, type = 'l', xlab = 'Time (d)', 
     ylab = expression('CH'[4]~'emission (g/d)'))
lines(CH4_emis_rate ~ time, data = out2, type = 'l', col = 'red')
```

Alternatively, the special `add_pars` argument can be used to specify just those parameters (or individual parameter elements) that will be changed from their defaults.

```{r}
out2b <- abm(365, 1, add_pars = list(pH = 6))
```

These two approaches provide identical results:

```{r}
all.equal(out2, out2b)
```

Note that the `man_pars` name is not needed for the `add_pars` option.

Many arguments for the `abm()` function are named lists or vectors.
These arguments--or even one element within them--can still be specified using `add_pars`.
For example, to change only the VFA value for `conc_fresh` the following call provides a shortcut compared to specifying all elements within the `conc_fresh` vector (as in the `out2` example above).

```{r}
out3 <- abm(365, 1, add_pars = list(pH = 6, conc_fresh.VFA = 10))
```

This shortcut is referred to as the "par.element" approach in the documentation, and the `.` is a special character used to separate parameter (here, `conc_fresh`) and element (here, `VFA`) names.
(If desired, a different character can be set with the `par_key` argument.)

Of course, specifying all elements is always an option,

```{r}
out3b <- abm(365, 1, add_pars = list(pH = 6, conc_fresh = list(S2 = 0.0, SO4 = 0.2, 
                                                               TAN = 1.0, VFA = 10, 
                                                               Sp = 65, COD = 160)))
```

as is specifying a complete argument of parameters (as in `out2` above).

Setting arguments is explored further in the section on defining microbial groups below (Section 6).

# 6. Output options
By default, the `abm()` function returns a data frame with cumulative CH~4~ emission and other state variables, normalized in a variety of ways.
In total there are more than 300 columns---the first 20 are shown below.

```{r}
out1 <- abm(365, 1)
out1[365, 1:20]
```

Microbial biomass values (g COD) are present in the columns that directly follow time (d).
Emission of CH~4~ and CO~2~ are included as cumulative values (g), rates (g/d), and both types are also normalized by loading of COD, degradable COD (`dCOD`), and VS (based on either instantaneous rates or cumulative values).
The fraction of loaded COD converted through methanogenesis, respiration, and sulfate reduction is also given--these variables start with `f_`.
For example, fractional conversion of COD to CH~4~ based on instantaneous rates and cumulative values are shown in the plot below.

```{r}
plot(f_COD_CH4_rate ~ time, data = out1, type = 'l', col = 'blue', ylab = 'COD conversion to CH4, rate')
lines(f_COD_CH4_cum ~ time, data = out1, col = 'red')
```

Overall results can be extracted by changing the `value` argument to `sum` (for summary).

```{r}
out1s <- abm(365, 1, value = 'sum')
out1s
```

And an arbitrary startup period can be excluded from these summary results using the `startup` argument.
For example, the first 100 days are excluded in the example below.

```{r}
out1s <- abm(365, 1, value = 'sum', startup = 100)
out1s
```

Alternatively, set the `value` argument to `'all'` for time series data and the summary.

# 7. Defining microbial groups
By default, the ABM model includes five methanogen groups.
Sulfate reducer groups can be included as well.
Each microbial group is characterized by 14 parameters that describe the rate of metabolism, biomass yield, decay rate, and the response to temperature, pH, and ammonia.
Additionally, values are needed for biomass concentrations in fresh slurry and the storage.
Lastly, an enrichment factor parameter is specified for each group (default of 0 means no enrichment).
Unlike VFA consumption, the rate of the combined hydrolysis and fermentation step is controlled by a simple temperature-dependent first-order rate constant.
Aerobic respiration is controlled by the mass transfer rate of O~2~ to the slurry surface.
For these two processes then, there is no (explicit) associated microbial group.

A central feature of the ABM model is the ability to specify any number of methanogenic groups. 
The default set of groups can be seen in the default values for the `grp_pars` argument.

```

```

To define a custom set, a single (albeit complex) argument `grp_pars` needs to be set. 
This task is straightforward if tedious (and is described later in this section).
A more common need is to tweak default parameters, which can be done efficiently using the `add_pars` argument (for **add**itional **par**rameter**s**).
For example, to increase `qhat_optim` of group `m1` to 3 g/g-d (g substrate COD per g biomass COD per day) and the yield to 0.06 g/g, the following call could be used:

```{r}
out4 <- abm(365, 1, add_pars = list(qhat_opt.m1 = 3, yield.m1 = 0.06))
plot(m1_conc ~ time, data = out4, type = 'l', ylab = 'Biomass concentration (g/kg)')
lines(m1_conc ~ time, data = out1, type = 'l', col = 'red')
```

This change from the default values (2.77 and 0.04) has a drastic effect, which perhaps should not be too surprising because the yield change alone represents a 50% improvement in fitness.

The `all` keyword can be used in `add_pars` to make changes to all microbial groups. 
For example, to completely turn off microbial decay, use the following call.


```{r}
out4a <- abm(365, 1, add_pars = list(decay_rate.all = 0))
matplot(out1$time, out1[, nn <- c('m1', 'm2', 'm3', 'm4', 'm5')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Microbial biomass (g)')
matlines(out4a$time, out4a[, nn <- c('m1', 'm2', 'm3', 'm4', 'm5')], lty = 2)
legend('topleft', nn, col = 1:6, lty = 1)
```

To define completely new microbial groups, the example of the default argument in the help file can be followed.
The special keywords `all` and `default` make the task easier.
All groups must be named individually in the `grps` parameter, which identifies the groups.
For any numeric element, use `all` if all groups have the same value (as in `xa_init` below).
Or, use `default` if some (unspecified) groups have the same value (as in `xa_fresh` below).

```{r}
grp_pars <- list(grps = c('m1', 'm2', 'm3', 'p1', 'p2', 'sr1'),
                 yield = c(default = 0.04, sr1 = 0.065),
                 xa_fresh = c(default = 0.001, sr1 = 0.001),
                 xa_init = c(all = 0.01), 
                 decay_rate = c(all = 0.02), 
                 ks_coefficient = c(m1 = 0.5, m2 = 1.5, m3 = 1.0, p1 = 1.0, p2 = 1.0, sr1 = 0.4),
                 resid_enrich = c(all = 0),
                 qhat_opt = c(m1 = 8, m2 = 13.33, m3 = 5.75, p1 = 2.77, p2 = 0.72, sr1 = 8.3),    
                 T_opt = c(m3 = 303, p1 = 293, p2 = 283, default = 313), 
                 T_min = c(m1 = 295.31, m2 = 295.31, m3 = 285.31, p1 = 275.31, p2 = 265.31, sr1 = 273),
                 T_max = c(m1 = 320.67, m2 = 320.67, m3 = 310.67, p1 = 300.67, p2 = 290.67, sr1 = 320.67),
                 ki_NH3_min = c(m1 = 0.01, m2 = 0.015, m3 = 0.015, p1 = 0.015, p2 = 0.015, sr1 = 0.015),
                 ki_NH3_max = c(m1 = 0.10, m2 = 0.131, m3 = 0.131, p1 = 0.131, p2 = 0.131, sr1 = 0.131),
                 ki_NH4_min = c(m1 = 1.70, m2 = 2.714, m3 = 2.714, p1 = 2.714, p2 = 2.714, sr1 = 2.714),
                 ki_NH4_max = c(m1 = 3.10, m2 = 4.764, m3 = 4.764, p1 = 4.764, p2 = 4.764, sr1 = 4.764),
                 pH_upr = c(m1 = 8.0, m2 = 8.0, m3 = 8.0, p1 = 8.0, p2 = 8.0, sr1 = 8.0),
                 pH_lwr = c(m1 = 6.5, m2 = 6.0, m3 = 6.5, p1 = 6.5, p2 = 6.5, sr1 = 6.0))

out5 <- abm(365, 1, grp_pars = grp_pars)
```

If some groups should be removed, this can be done easily with the `grps` element.
For example, to limit the groups to only `m1` and `m4`, use the following call.
Sulfate reducers are not required.
And a single group can be used.

```{r}
out4b <- abm(365, 1, add_pars = list(grps = c('m1', 'm4')))
head(out4b, 2)
plot(m4_conc ~ time, data = out4b, type = 'l', ylab = 'Biomass concentration (g/kg)')
lines(m4_conc ~ time, data = out1, type = 'l', col = 'red')

out4c <- abm(365, 1, add_pars = list(grps = 'm1'))
```

# 8. Simulating reactors
The ABM model inherently describes a reactor with continuous feeding and intermittent wasting.
To approximate a continuous reactor (which is not actually "continuous" in practice but typically has intermittent feeding and wasting--but this is a seprate discussion) the `resid_frac` argument can be set to a high value, e.g. `0.95`.
This provides frequent wasting of a small quantity. 
The following example simulates the startup of a mesophilic completely mixed anaerobic digester fed cattle manure (based on defaults).

```{r}
out5 <- abm(365, 1, add_pars = list(temp_C = 35, resid_frac = 0.95, alpha_opt = 0.2, 
                                    slurry_mass = 0.95 * 33333, slurry_prod_rate = 500))
```

Due to the structure of the code (the ODE solver is called separately for each filling interval), a drawback of this high `resid_frac` approach is a long evaluation time.

```{r}
plot(CH4_emis_cum_VS ~ time, data = out5, type = 'l')
plot(22300 / 16 * CH4_emis_cum_VS ~ time, data = out5, type = 'l')
plot(f_COD_CH4_rate ~ time, data = out5, type = 'l')

matplot(out5$time, out5[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1)
legend('topleft', nn, col = 1:5, lty = 1)

matplot(out5$time, out5[, c('Sp_conc', 'VFA_conc', 'COD_conc', 'dCOD_conc')], 
        type = 'l', lty = 1)
```

Batch reactors can also be simulated.

```{r}
out5b <- abm(365, 1, add_pars = list(temp_C = 35, alpha_opt = 0.2, 
                                    slurry_mass = 33333, slurry_prod_rate = 0))

```

```{r}
plot(CH4_emis_cum ~ time, data = out5b, type = 'l')
plot(22300 / 16 * CH4_emis_cum ~ time, data = out5b, type = 'l')

matplot(out5b$time, out5b[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1)
legend('topleft', nn, col = 1:5, lty = 1)

matplot(out5b$time, out5b[, c('Sp_conc', 'VFA_conc', 'COD_conc', 'dCOD_conc')], 
        type = 'l', lty = 1)
```

# 9. Variable temperature
Predicting short- and long-term responses to temperature change was a central objective of the ABM model.
Variable temperature is entered in a data frame with two columns.
For example, gradual warming from 10&deg;C to 25&deg;C, a hold, and then a gradual cooling back to 10&deg;C can be specified as shown in the `temp_dat` data frame constructed below.

```{r}
temp_dat <- data.frame(time = 100 + c(0, 60, 220, 280), 
                       temp_C =     c(10, 25,  25,   10)) 
plot(temp_C ~ time, data = temp_dat, type = 'l')
```

The model can either interpolate (the default) or use constant temperatures between change points. 
The temperature data can be supplied using the `mng_pars` argument or, more simply, with `add_pars`.

```{r}
out6 <- abm(500, 1, add_pars = list(temp_C = temp_dat))
```

```{r, echo = FALSE}
plot(temp_C ~ time, type = 'l', col = 'red', data = out6)
abline(v = temp_dat$time, lty = 2, col = 'gray45')

matplot(out6$time, out6[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Biomass conc. (g/kg)')
legend('topleft', nn, col = 1:5, lty = 1)
abline(v = temp_dat$time, lty = 2, col = 'gray45')

plot(CH4_emis_rate_VS ~ time, data = out6, type = 'l')
abline(v = temp_dat$time, lty = 2, col = 'gray45')

plot(VFA_conc ~ time, type = 'l', col = 'purple', data = out6)
abline(v = temp_dat$time, lty = 2, col = 'gray45')
```

NTS: Really no development of a psychrophilic population?
NTS: What does it take?

Temperature data could be much higher resolution, e.g., daily values.

For anaerobic digesters, the effect of both short- and long-term changes are of interest.
In a controlled environment, temperature change is not always gradual, but can be (deliberately) rapid.
The `approx_method_temp` argument can be used for this type of pattern, instead of the linear interpolation shown above (which is the default).
The following data frame can be used to simulate a reactor initially running at 35&deg;C suddenly reduced to 25&deg;C for 5 days, followed by stabilization and finally a much longer temperature change.

```{r}
temp_dat <- data.frame(time = 300 + c(0,  1,  6, 100), 
                       temp_C =    c(35, 25, 35,  25)) 
plot(temp_C ~ time, data = temp_dat, type = 's')
```

```{r}
out7 <- abm(600, 1, add_pars = list(temp_C = temp_dat, resid_frac = 0.95, 
                                    slurry_mass = 0.95 * 33333, slurry_prod_rate = 500),
            approx_method_temp = 'constant')

```

```{r}
plot(temp_C ~ time, type = 'l', col = 'red', data = out7)
abline(v = temp_dat$time, lty = 2, col = 'gray45')

matplot(out7$time, out7[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Biomass conc. (g/kg)')
legend('topleft', nn, col = 1:5, lty = 1)
abline(v = temp_dat$time, lty = 2, col = 'gray45')

plot(CH4_emis_rate_VS ~ time, data = out7, type = 'l')
abline(v = temp_dat$time, lty = 2, col = 'gray45')

plot(VFA_conc ~ time, type = 'l', col = 'purple', data = out7)
abline(v = temp_dat$time, lty = 2, col = 'gray45')
```

# 10. Acidification
Acidification of slurry with sulfuric acid (H~2~SO~4~) is an effective approach for reducing CH~4~ emission.
In the ABM model, there are a few options for how acidification can be specified.
With data on slurry pH vs. time, the `pH` parameter can be used.
As with temperature, both instant changes and linear interpolation can be used.
The following data might come from a channel where acidification took effect over 30 days, was used for 100, and then stopped.

```{r}
pH_dat <- data.frame(time = c(200, 230, 330, 360), pH = c(7.5, 5.0, 5.0, 7.5))
plot(pH ~ time, data = pH_dat, type = 'l')
```

As with temperature, only the times that pH changes are needed--earlier and later times extend the nearest value.

```{r}
out8 <- abm(730, 1, add_pars = list(pH = pH_dat), approx_method_pH = 'linear')
```

```{r}
plot(pH ~ time, type = 'l', col = 'red', data = out8)
abline(v = pH_dat$time, lty = 2, col = 'gray45')

matplot(out8$time, out8[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Biomass conc. (g/kg)')
legend('topleft', nn, col = 1:5, lty = 1)
abline(v = pH_dat$time, lty = 2, col = 'gray45')

plot(CH4_emis_rate_VS ~ time, data = out8, type = 'l')
abline(v = pH_dat$time, lty = 2, col = 'gray45')

plot(VFA_conc ~ time, type = 'l', col = 'purple', data = out8)
abline(v = pH_dat$time, lty = 2, col = 'gray45')
```

Alternatively, pH can be calculated from the H~2~SO~4~ concentration (given as SO~4~^-2^).
This approach is based on a titration curve for "typical" slurry.

```{r}
out9 <- abm(200, 1, add_pars = list(conc_fresh.SO4 = 1.3, pH = 'calc'))
```

```{r}
plot(pH ~ time, type = 'l', col = 'red', data = out9)
abline(v = pH_dat$time, lty = 2, col = 'gray45')

matplot(out9$time, out9[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Biomass conc. (g/kg)')
legend('topleft', nn, col = 1:5, lty = 1)
abline(v = pH_dat$time, lty = 2, col = 'gray45')

plot(CH4_emis_rate_VS ~ time, data = out9, type = 'l')
abline(v = pH_dat$time, lty = 2, col = 'gray45')

plot(VFA_conc ~ time, type = 'l', col = 'purple', data = out9)
abline(v = pH_dat$time, lty = 2, col = 'gray45')
```

NTS: This is a difficult approach and the titration curve needs another look.

# 11. Model output as input
Although the `abm()` function is quite flexible and can accommodate changes in inputs over time, there is plenty it cannot do.
To partially address limitations, it is possible to specify output from one call as the starting conditions for a new call.
This is done using the `starting` argument.
In the following example, two calls are used to show the effect of an instant change in the substrate concentration in fresh slurry.

```{r}
out10 <- abm(365, 1, add_pars = list(conc_fresh.Sp = 100, conc_fresh.COD = 200, conc_fresh.VFA = 10))
```

```{r}
out11 <- abm(365, 1, add_pars = list(conc_fresh.Sp = 10, conc_fresh.COD = 20, conc_fresh.VFA = 0),
             starting = out10)
```

```{r}
plot(Sp_conc ~ time, type = 'l', col = 'purple', data = out11)

matplot(out11$time, out11[, nn <- c('m1_conc', 'm2_conc', 'm3_conc', 'm4_conc', 'm5_conc')], 
        type = 'l', lty = 1, xlab = 'Time (d)', ylab = 'Biomass conc. (g/kg)')
legend('topleft', nn, col = 1:5, lty = 1)

plot(CH4_emis_rate_VS ~ time, data = out11, type = 'l')

plot(VFA_conc ~ time, type = 'l', col = 'purple', data = out11)
```

# 12. More information
Users can track development of the ABM package on GitHub: [https://github.com/sashahafner/ABM](https://github.com/sashahafner/ABM).
To report bugs or request features, use the "Issues" page.
More details will be available in publications in the near future.

